\chapter{Brownian Motion}

Brownian motion is the canonical model for \emph{random continuous motion}.
It appears (i) as a mathematical idealization of the irregular movement of particles
suspended in a fluid, (ii) as a universal scaling limit of many random-walk models,
and (iii) as the driving noise in stochastic differential equations.

In these notes we work on the time interval $[0,1]$ for simplicity. (Once existence is
established on $[0,1]$, the extension to $[0,\infty)$ can be obtained by concatenation
of independent copies, or by standard projective/extension arguments.)

\begin{definition}[Standard Brownian motion on $[0,1]$]
A stochastic process $B=(B_t)_{t\in[0,1]}$ on a probability space
$(\Omega,\mathcal F,\Prob)$ is called a \emph{standard Brownian motion} if
\begin{enumerate}
\item $B_0=0$ almost surely,
\item $t\mapsto B_t(\omega)$ is continuous on $[0,1]$ for $\Prob$-almost every $\omega$,
\item for all $0\le s<t\le 1$,
\[
B_t-B_s \sim \mathcal N(0,t-s),
\]
\item increments over disjoint intervals are independent.
(Equivalently: for any $0\le t_0<t_1<\dots<t_m\le 1$, the increments
$B_{t_1}-B_{t_0},\dots,B_{t_m}-B_{t_{m-1}}$ are independent.)
\end{enumerate}
\end{definition}

Intuitively:
\emph{stationary increments} means the law of $B_t-B_s$ depends only on $t-s$,
\emph{independent increments} means there is no memory across disjoint time intervals,
and \emph{continuity} rules out jumps.

\section{Properties of the standard Brownian motion}
A key fact (and the entry point for the construction in the next chapter) is that
Brownian motion is a centered Gaussian process with covariance $\min(s,t)$.

\begin{proposition}[Mean and covariance]
Let $B$ be a Brownian motion on $[0,1]$. Then for all $s,t\in[0,1]$,
\[
\E[B_t]=0,
\qquad
\E[B_sB_t]=\min(s,t).
\]
\end{proposition}

\begin{proof}
First we show that $\E[B_t]=0$ for all $t\in[0,1]$.
Indeed, for $0\le s<t\le 1$ we have $B_t-B_s\sim\mathcal N(0,t-s)$, hence
$\E[B_t-B_s]=0$ and therefore $\E[B_t]=\E[B_s]$.
Thus $t\mapsto \E[B_t]$ is constant, and since $B_0=0$ a.s.\ we get
$\E[B_t]=\E[B_0]=0$.


Next, we show that $\E[B_sB_t]=\min(s,t)$. Fix $0\le s\le t\le 1$. By the increment decomposition,
\[
B_t = B_s + (B_t-B_s),
\]
and by definition the increment $B_t-B_s$ is independent of $B_s$ and has mean $0$.
Therefore
\[
\E[B_sB_t]
=\E\!\big[B_s(B_s+(B_t-B_s))\big]
=\E[B_s^2] + \E[B_s]\E[B_t-B_s]
=\E[B_s^2].
\]
But $B_s\sim\mathcal N(0,s)$, hence $\E[B_s^2]=\Var(B_s)=s=\min(s,t)$.
\end{proof}

\begin{proposition}[Conversely: covariance $\min(s,t)$ determines Brownian motion]
Let $B=(B_t)_{t\in[0,1]}$ be a \emph{centered Gaussian process} such that $B_0=0$ a.s.,
$t\mapsto B_t(\omega)$ is continuous for $\Prob$-a.e.\ $\omega$, and
\[
\E[B_sB_t]=\min(s,t)\qquad (s,t\in[0,1]).
\]
Then $B$ is a standard Brownian motion on $[0,1]$.
\end{proposition}

\begin{proof}[Proof sketch]
For $0\le s<t\le 1$,
\[
\Var(B_t-B_s)
=\E[(B_t-B_s)^2]
=\E[B_t^2]+\E[B_s^2]-2\E[B_sB_t]
=t+s-2s=t-s,
\]
so $B_t-B_s\sim\mathcal N(0,t-s)$ (Gaussianity is preserved under linear maps).
For disjoint intervals $[s,t]$ and $[u,v]$ with $t\le u$, compute
\[
\Cov(B_t-B_s,\;B_v-B_u)
=\min(t,v)-\min(t,u)-\min(s,v)+\min(s,u)=0.
\]
Because the pair of increments is jointly Gaussian, zero covariance implies
independence. This gives independent increments. Continuity and $B_0=0$ are assumed,
hence $B$ is Brownian motion.
\end{proof}

\paragraph{Marginals and finite-dimensional distributions}
From the definition (or from the covariance characterization) we immediately have
\[
B_t \sim \mathcal N(0,t),\qquad t\in[0,1].
\]
More generally, for any $0\le t_1<\dots<t_m\le 1$, the vector
$(B_{t_1},\dots,B_{t_m})$ is multivariate normal, since it can be written as a linear
transform of the independent Gaussian increments
\[
\Delta_j := B_{t_j}-B_{t_{j-1}},\qquad t_0:=0.
\]

\paragraph{Scaling and symmetry}
For $c>0$, the rescaled process
\[
\widetilde B_t := \frac{1}{\sqrt c}\, B_{ct},
\qquad t\in\Big[0,\frac{1}{c}\Big],
\]
is again a standard Brownian motion (on its time domain). Also, $(-B_t)_{t\in[0,1]}$
has the same law as $(B_t)_{t\in[0,1]}$.

\paragraph{Regularity}
Brownian paths are continuous but extremely rough: they are almost surely nowhere
differentiable. A useful quantitative statement (proved later in many courses, e.g.
via Kolmogorov--Chentsov or directly from the series construction) is that for every
$\alpha\in(0,\tfrac12)$ the paths are $\alpha$-H\"older continuous on $[0,1]$
almost surely, while no exponent $\alpha\ge \tfrac12$ is possible almost surely.

\section{Construction of Brownian motion}
The definition of Brownian motion postulates the existence of a process with very
specific properties. Standard constructions include:
\begin{itemize}
\item Kolmogorov extension (construct consistent finite-dimensional distributions) +
      a continuity theorem (obtain a continuous modification),
\item the L\'evy--Ciesielski (Haar--Schauder) construction: an explicit random series
      built from a complete orthonormal system of $L^2([0,1])$,
\item invariance principles (Brownian motion as a scaling limit of random walks).
\end{itemize}
We now construct a centered Gaussian process with covariance $\min(s,t)$ and continuous paths, and then identify it as Brownian motion.
Below we follow the Haar--Schauder approach, since it gives an explicit
realization \emph{and} a natural approximation scheme by truncating the series.


Brownian motion can be characterized as a centered Gaussian process whose covariance
structure is given by
\[
\E[B(s)B(t)] = \min(s,t), \qquad s,t \in [0,1].
\]
The guiding principle behind the construction below is to \emph{build such a Gaussian
process explicitly} by expanding the indicator functions
$1_{[0,t]}$ in an orthonormal basis of $L^2([0,1])$.

Indeed, if $\{g_{k,n}\}$ is an orthonormal basis of $L^2([0,1])$, then any function
$h \in L^2([0,1])$ admits the expansion
\[
h = \sum_{k,n} \langle h, g_{k,n} \rangle g_{k,n},
\]
and Parseval's identity ensures that inner products are preserved by summation over the
basis. Applying this idea formally to the family $\{1_{[0,t]}\}_{t\in[0,1]}$, one is led
to consider random series of the form
\[
B(t) = \sum_{k,n} Z_{k,n} \int_0^t g_{k,n}(u)\,du,
\]
where $\{Z_{k,n}\}$ are independent standard normal random variables.

The choice of the Haar basis is particularly natural: its elements are localized on
dyadic intervals, and their integrals are continuous, piecewise linear ``tent''
functions. This localization allows for precise control of the uniform convergence of
the series and ultimately guarantees the continuity of the sample paths.

The proof below follows this strategy rigorously. First, we show that the random series
converges uniformly almost surely, producing a continuous process. Then we identify the
finite-dimensional distributions and compute the covariance via Parseval's identity.
The Gaussian structure then automatically yields independent, stationary increments,
revealing the constructed process to be Brownian motion.

\begin{theorem}[Existence (L\'evy--Ciesielski / Haar--Schauder Construction)]
There exists a probability space $(\Omega,\mathcal F,\Prob)$ and a stochastic process
$B=(B_t)_{t\in[0,1]}$ such that

\begin{enumerate}
\item $B_0=0$ almost surely,
\item $t\mapsto B_t(\omega)$ is continuous on $[0,1]$ for $\Prob$-almost every $\omega$,
\item for all $0\le s<t\le 1$, $B_t-B_s\sim\mathcal N(0,t-s)$,
\item increments over disjoint intervals are independent.
\end{enumerate}

Hence $B$ is a standard Brownian motion on $[0,1]$.

Moreover, $B$ admits the series representation
\[
B(t)=\sum_{n=0}^\infty\sum_{k=1}^{2^n} Z_{k,n}\, f_{k,n}(t),
\]
where $\{Z_{k,n}\}$ are independent standard normal random variables and
$f_{k,n}(t)=\int_0^t g_{k,n}(u)\,du$, with $\{g_{k,n}\}$ the Haar orthonormal basis of
$L^2([0,1])$.
\end{theorem}

\begin{proof}

The proof proceeds in four steps:
(i) definition of the deterministic basis,
(ii) almost sure uniform convergence of the random series,
(iii) identification of the Gaussian covariance,
(iv) verification of the Brownian motion properties.

\Step{1: Haar functions and Schauder primitives}

Let $\{g_{k,n}\}_{n\ge0,\,1\le k\le2^n}$ be the Haar orthonormal basis of $L^2([0,1])$,
defined by
\[
g_{k,n}(t)
=
2^{n/2}\big(1_{I_{k,n}^\ell}(t)-1_{I_{k,n}^r}(t)\big),
\]
where $I_{k,n}=[(k-1)2^{-n},k2^{-n})$ and $I_{k,n}^\ell$, $I_{k,n}^r$ are its left and right
halves.

Define the integrated (Schauder) functions
\[
f_{k,n}(t)=\int_0^t g_{k,n}(u)\,du.
\]
Each $f_{k,n}$ is continuous, piecewise linear, and supported on $I_{k,n}$.
A direct computation shows
\[
\|f_{k,n}\|_\infty \le 2^{-(n/2+1)}.
\]

\Step{2: Definition of the random series}

Let $\{Z_{k,n}\}$ be independent $\mathcal N(0,1)$ random variables.
For $N\in\mathbb{N}$ define the partial sums
\[
B_N(t)=\sum_{n=0}^N\sum_{k=1}^{2^n} Z_{k,n} f_{k,n}(t).
\]
Each $B_N$ is a continuous stochastic process on $[0,1]$.

\Step{3: Almost sure uniform convergence}

Define the level-$n$ contribution
\[
\Delta_n(t)=\sum_{k=1}^{2^n} Z_{k,n} f_{k,n}(t).
\]
Since the supports of $f_{k,n}$ are disjoint for fixed $n$,
\[
\sup_{t\in[0,1]}|\Delta_n(t)|
\le \max_{1\le k\le2^n}|Z_{k,n}|\cdot \|f_{k,n}\|_\infty
\le 2^{-(n/2+1)} \max_{k}|Z_{k,n}|.
\]

Using Gaussian tail estimates and the union bound, one shows that
\[
\sum_{n=0}^\infty \Prob\!\left(\max_{k}|Z_{k,n}| > a_n\right)<\infty
\]
for a suitable sequence $a_n\to\infty$.
By the Borel--Cantelli lemma,
\[
\sum_{n=0}^\infty \sup_{t\in[0,1]}|\Delta_n(t)| < \infty
\quad\text{almost surely}.
\]
Hence $\sum_n \Delta_n(t)$ converges uniformly in $t$ almost surely.
Define
\[
B(t)=\sum_{n=0}^\infty \Delta_n(t).
\]
Then $B$ has almost surely continuous sample paths.

\Step{4: Gaussianity and covariance structure}

For each fixed $t$, $B_N(t)$ is a finite linear combination of independent Gaussian
random variables, hence Gaussian. Moreover, $B_N(t)\to B(t)$ in $L^2$, so $B(t)$ is
Gaussian.

For $s,t\in[0,1]$,
\[
\E[B(s)B(t)]
=
\sum_{n,k} f_{k,n}(s)f_{k,n}(t).
\]
Since
\[
f_{k,n}(t)=\int_0^t g_{k,n}(u)\,du
= \langle 1_{[0,t]}, g_{k,n}\rangle_{L^2},
\]
the bilinear Parseval identity yields
\[
\E[B(s)B(t)]
=
\langle 1_{[0,s]},1_{[0,t]}\rangle_{L^2}
=
\int_0^{\min(s,t)}1\,du
=
\min(s,t).
\]

\Step{5: Identification as Brownian motion}

Since $(B(s),B(t))$ is jointly Gaussian,
\[
B(t)-B(s)\sim\mathcal N(0,t-s).
\]
Moreover, for disjoint intervals $[s,t]$ and $[u,v]$,
\[
\mathrm{Cov}(B(t)-B(s),\,B(v)-B(u))=0,
\]
which implies independence of increments.
Finally, $B_0=0$ holds trivially from the definition.

Therefore $B$ is a standard Brownian motion on $[0,1]$.

\end{proof}

\begin{remark}
This construction provides both a theoretical realization of Brownian motion
and a practical approximation scheme: truncating the series yields continuous
random paths converging uniformly almost surely.
\end{remark}

\begin{figure}[h]
\centering
\begin{tikzpicture}[x=11cm,y=1.2cm,>=Latex]

% Parameters as labels (we draw generically; coordinates are symbolic on the axis)
% We'll place the key points at 0.20, 0.55, 0.90 just for the sketch geometry:
\def\a{0.20} % (k-1)2^{-n}
\def\m{0.55} % (k-1/2)2^{-n}
\def\b{0.90} % k2^{-n}

% axes
\draw[->] (0,0) -- (1.05,0) node[below] {$t$};
\draw[->] (0,-1.15) -- (0,1.15);

% y-level marks (symbolic)
\draw (-0.02,1) -- (0.02,1) node[left] {$2^{n/2}$};
\draw (-0.02,-1) -- (0.02,-1) node[left] {$-2^{n/2}$};

% Haar function g_{k,n}
% zero part (optional faint guide)
\draw[gray!40] (0,0) -- (\a,0);
\draw[gray!40] (\b,0) -- (1,0);

% left half: +2^{n/2}
\draw[thick] (\a,0) -- (\a,1);
\draw[thick] (\a,1) -- (\m,1);

% jump at midpoint
\draw[thick] (\m,1) -- (\m,-1);

% right half: -2^{n/2}
\draw[thick] (\m,-1) -- (\b,-1);
\draw[thick] (\b,-1) -- (\b,0);

% x-ticks and labels
\draw (\a,0.04) -- (\a,-0.04) node[below=4pt] {$(k-1)2^{-n}$};
\draw (\m,0.04) -- (\m,-0.04) node[below=4pt] {$(k-\tfrac12)2^{-n}$};
\draw (\b,0.04) -- (\b,-0.04) node[below=4pt] {$k2^{-n}$};

% braces / interval labels (optional)
\node[above] at ({(\a+\m)/2},1) {$I_{k,n}^\ell$};
\node[below] at ({(\m+\b)/2},-1) {$I_{k,n}^r$};

\node at (0.53,1.05) {$g_{k,n}$};

\end{tikzpicture}
\caption{Sketch of the Haar function $g_{k,n}(t)=2^{n/2}(1_{I^\ell_{k,n}}(t)-1_{I^r_{k,n}}(t))$.}
\end{figure}


\begin{figure}[h]
\centering
\begin{tikzpicture}[x=11cm,y=2.2cm,>=Latex]

% Place the endpoints for the sketch geometry (symbolic labels on ticks)
\def\a{0.20} % a=(k-1)2^{-n}
\def\m{0.55} % m=(k-1/2)2^{-n}
\def\b{0.90} % b=k2^{-n}

% Height corresponds to ||f_{k,n}||_\infty = 2^{-n/2-1}
\def\H{0.55}

% axes
\draw[->] (0,0) -- (1.05,0) node[below] {$t$};
\draw[->] (0,0) -- (0,0.85);

% peak height tick
\draw (-0.02,\H) -- (0.02,\H) node[left] {$2^{-n/2-1}$};

% baseline faint
\draw[gray!40] (0,0) -- (\a,0);
\draw[gray!40] (\b,0) -- (1,0);

% tent function f_{k,n}
\draw[thick] (\a,0) -- (\m,\H) -- (\b,0);

% x-ticks and labels
\draw (\a,0.02) -- (\a,-0.02) node[below=4pt] {$(k-1)2^{-n}$};
\draw (\m,0.02) -- (\m,-0.02) node[below=4pt] {$(k-\tfrac12)2^{-n}$};
\draw (\b,0.02) -- (\b,-0.02) node[below=4pt] {$k2^{-n}$};

% annotations
\node[above] at ({(\a+\m)/2},{\H/2}) {$\text{slope }=2^{n/2}$};
\node[above] at ({(\m+\b)/2},{\H/2}) {$\text{slope }=-2^{n/2}$};
\node[above] at (\m,\H) {$f_{k,n}(m)=2^{-n/2-1}$};

\node at (0.56,0.80) {$f_{k,n}$};

\end{tikzpicture}
\caption{Sketch of the Schauder function $f_{k,n}(t)=\int_0^t g_{k,n}(u)\,du$ (a continuous ``tent'' supported on $I_{k,n}$).}
\end{figure}


